---
title: "When interpreting plots, LLMs see what they want to see: Part 2"
format:
  html:
    toc: true
    toc-depth: 2
    echo: false
    warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(bluffbench)

# Load consolidated blog data
blog_data <- read_rds(here::here("data/blog_data.rds"))
baseline_reading <- blog_data$baseline_reading
intervention_results <- blog_data$intervention_results
mitm_helper <- blog_data$mitm_helper
```

```{r theme-setup}
#| include: false
theme_set(
  theme_bw(base_size = 14) +
    theme(
      panel.border = element_blank(),
      plot.subtitle = element_text(face = "italic"),
      legend.position = "bottom",
      strip.background = element_rect(fill = "#f0f0f0")
    )
)

colors <- c("Correct" = "#67a9cf", "Incorrect" = "#ef8a62")

# Helper for heatmaps
plot_heatmap <- function(data, facet_var = NULL) {
  p <- data |>
    mutate(id = str_replace_all(id, "_", " ") |> str_wrap(20)) |>
    ggplot(aes(x = factor(epoch), y = id, fill = score)) +
    geom_tile(color = "white", linewidth = 0.5) +
    scale_fill_manual(values = colors) +
    labs(x = "Epoch", y = NULL, fill = NULL)

  if (!is.null(facet_var)) {
    p <- p + facet_wrap(vars({{ facet_var }}), ncol = 2)
  }
  p
}
```

In [a previous post](https://posit.co/blog/introducing-bluffbench/), we discussed the results from [bluffbench](https://github.com/simonpcouch/bluffbench), which evaluates how well LLMs can interpret plots that contradict with their prior beliefs. The models tested (GPT-5, Claude Sonnet 4.5, and Gemini Pro 2.5) largely failed to accurately interpret such plots, especially when plotting known datasets (`diamonds`, `mpg`, etc.) which had been secretly manipulated. 

For example, one of the samples inverted the price-carat relationship in `diamonds` so that larger diamonds are less expensive. 

```{r diamonds-comparison}
#| fig-width: 9
#| fig-height: 4
#| fig-alt: "Two scatterplots side by side. Left: original diamonds data showing positive relationship between carat and price. Right: manipulated data showing inverted (negative) relationship."
bind_rows(
  diamonds |>
    select(carat, price) |>
    mutate(version = "Original data"),
  diamonds |>
    mutate(price = max(price) - price) |>
    select(carat, price) |>
    mutate(version = "Manipulated data")
) |>
  mutate(version = fct_inorder(version)) |>
  ggplot(aes(x = carat, y = price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  facet_wrap(vars(version)) +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    x = "Carat",
    y = "Price",
    title = "Manipulated diamonds data from bluffbench",
    subtitle = "Models plot the manipulated data, but typically report seeing the expected relationship"
  )
```

None of the models ever accurately reported seeing this negative relationship. Instead, they reported that the plot showed the expected, positive relationship. Here's a portion of a response from Claude Sonnet 4.5:

> Here's the plot showing the relationship between diamond carat and price! How diamond size affects price: 
>
> There's a strong positive relationship between carat (diamond size) and price.^[You can see the full log for this sample [here](https://simonpcouch.github.io/bluffbench/logs/#/logs/2025-10-21T13-47-37-05-00_bluffbench_465475d21aad6cc3ab920b.json/samples/sample/diamonds_carat_price_reversal/1/).]

We ended that post by saying that we were still looking for a solution to the problem, which is important because data analysis agents like Databot rely on plot interpretation to understand data. Although we still haven't solved the problem, we do have some updates. 

## What's really going on?

First, _why_ exactly do LLMs misinterpet plots like those in bluffbench?

In our previous post, we showed the results from the two adversarial conditions (mocked and intuitive), as well as a baseline condition, for three different LLMs. 

```{r bluffbench-all}
#| fig-width: 9
#| fig-height: 5
#| fig-alt: "Grouped bar chart showing model accuracy across bluffbench conditions. All models perform worse on mocked datasets."
bluff_results |>
  mutate(
    type = fct_relevel(type, "baseline", "intuitive", "mocked") |>
      fct_recode(
        "Baseline" = "baseline",
        "Intuitive" = "intuitive",
        "Mocked" = "mocked"
      )
  ) |>
  summarize(
    pct_correct = mean(score == "C"),
    .by = c(model, type)
  ) |>
  ggplot(aes(y = type, x = pct_correct, fill = model)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_viridis_d(option = "plasma", begin = 0.2, end = 0.8) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    x = "Percent correct",
    y = NULL,
    fill = NULL,
    title = "Bluffbench results for mocked, intuitive, and baseline conditions",
    subtitle = "Models perform worse when the data directly conradicts their priors (mocked condition)"
  )
``` 

The models perform moderately on the baseline, but still perform worse than you might expect a frontier LLM, especially one you're using to power a data science agent, to do.^[One contributing factor is that models in this baseline condition occasionally created their own dataset to plot rather than using the one provided, meaning they weren't even looking at the correct data. We chose not to prevent this behavior since we wanted to simulate how models might behave in similar real-world tasks.]

<!-- insert plot from bluffbench -->

This leads to one possible hypothesis: maybe the current LLMs actually can't interpret plots very well. It's possible they have "blurry" vision when looking at the plot image or trouble attending to the actual patterns present in the plot. Their limited vision or attention may be enough to adequately (but not perfectly) interpret plots that align with their expectations, especially if they can rely on information like axis labels and dataset names, but severely limits them when plots are not what that expect (as in the mocked and intuitive bluffbench cases). 

To delineate it further, here are three possible hypotheses for what's going wrong:

1. **Skill issue:** Current LLMs just aren't very good at reading plots. They may appear competent in some situations by relying on other skills, like having prior knowledge of the data or incorporating other contextual information (data set name, axis labels, code, etc.).

2. **Attention issue:** Models _can_ read plots, but don't bother looking at them carefully when they think they already know the answer.

3. **Priors issue:** Models can read plots fine, but their expectations override what they actually see. These expectations may come into play even in seemingly benign cases like the bluffbench baseline scenarios. 

To tease apart these hypotheses, we ran two additional baseline evaluations. 

## LLMs can interpret plots accurately

We implemented two additional baseline evaluations:

- **Image only:** Models were shown just the plot image and asked to describe the pattern they observed. 
- **Code + image:** Models were instructed to write code to plot a particular dataset, inspect the result, and describe what they observed. 

Importantly, the visualizations created by the models and the ones provided to them displayed the same data. The datasets had generic names (e.g., `df`) and column names (`x`, `y`, etc.). 

We tested two scenarios with straightforward, non-adversarial plots:

```{r baseline-reading}
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Faceted bar chart showing model accuracy. Image only: 95-100% correct. Code + image: 70-90% correct."
baseline_reading |>
  mutate(
    task = fct_recode(
      task,
      "Model sees image only" = "Image only",
      "Model writes code, sees result" = "Code + image"
    ),
    task = fct_relevel(task, "Model sees image only")
  ) |>
  ggplot(aes(y = model, fill = score)) +
  geom_bar(position = "fill") +
  facet_wrap(vars(task)) +
  scale_fill_manual(breaks = rev, values = colors) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x = "Percent correct",
    y = NULL,
    fill = NULL,
    title = "Baseline plot interpretation",
    subtitle = "Models are accurate at interpreting straightforward plots"
  )
```

Interestingly, the models did _better_ when just shown the image, suggesting that even non-adversarial information could interfere or activate priors that interefered with the models ability to interpret plots.

This result suggests that model's difficulty with plot interpretation is fundamentally __not a skill issue.__ Their vision is fine. What they are bad at is incorporating the other information they have about the data or plot into their understanding of the actual visual patterns seen in the image. 

(for example, just knowing its a scatterplot seems to activate the models prior that it will be a linear plot)

This suggests taht it's either an attention issue, a priors issue, or possibly a combination. Regardless it's important that we know that models are actually fine at seeing the plots and visually interpreting them. This also rules out other possible issues, like the images being encoded in a way that makes them difficult for the models to interpret. 

## Interventions

After establishing this baseline, we continued to try several interventions, with only limited success. Here are a few fixes we tried:

**Memo prompt:** We asked models to first describe only the visual elements of the plot, as if they had no prior knowledge of the data and subject matter, and ignoring axis labels, and to put this description within `<MEMO>` tags. You can see the exact prompt here. 

**Extended thinking:** The memo prompt had some success, so we built on that prompt, but added language to prompt the models to use more extended thinking. 

Neither prompting technique, nor others we tried, had much success, so we tried a more structural intervention next. 

**Model-in-the-middle:** We had a fresh model (the _model in the middle_) describe the plot, then pass that text description to the primary model, which produced the final description of the plot. 

```{r intervention-comparison}
#| fig-width: 9
#| fig-height: 4
#| fig-alt: "Bar chart comparing intervention performance on mocked datasets. All interventions show high failure rates."
intervention_results |>
  filter(type == "mocked") |>
  mutate(
    intervention = fct_relevel(
      intervention,
      "Model-in-the-middle",
      "Thinking",
      "Memo prompt",
      "None"
    )
  ) |>
  ggplot(aes(y = intervention, fill = score)) +
  geom_bar(position = "fill") +
  scale_fill_manual(breaks = rev, values = colors) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x = "Percent correct",
    y = NULL,
    fill = NULL,
    title = "Interventions on mocked (adversarial) datasets",
    subtitle = "None of the interventions improved accuracy above 25%"
  )
```

<!--

```{r intervention-heatmap}
#| fig-width: 10
#| fig-height: 6
#| fig-alt: "Heatmap showing per-sample accuracy for each intervention across epochs. Most samples show consistent failures regardless of intervention."
#| eval: false
intervention_results |>
  filter(type == "mocked", intervention != "None") |>
  mutate(
    intervention = fct_relevel(
      intervention,
      "Memo prompt",
      "Thinking",
      "Model-in-the-middle"
    )
  ) |>
  (\(d) plot_heatmap(d, intervention))()
```

-->

## The model in the middle can see just fine

The model-in-the-middle interpreted plots accurately about 75% of the time. However, the primary models often doubts the model-in-the-middle's description, even when we explicitly told it to trust the description as ground truth.

```{r mitm-comparison}
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Bar chart comparing MITM helper accuracy (75%) versus primary model accuracy (15%) on mocked datasets."
mitm_helper_summary <- mitm_helper |>
  filter(type == "mocked") |>
  summarize(pct = mean(score == "Correct")) |>
  mutate(model = "MITM helper\n(no context)")

mitm_primary_summary <- intervention_results |>
  filter(type == "mocked", intervention == "Model-in-the-middle") |>
  summarize(pct = mean(score == "Correct")) |>
  mutate(model = "Primary model\n(with context)")

bind_rows(mitm_helper_summary, mitm_primary_summary) |>
  mutate(model = fct_inorder(model)) |>
  ggplot(aes(x = model, y = pct)) +
  geom_col(fill = "#67a9cf", width = 0.6) +
  geom_text(
    aes(label = scales::percent(pct, accuracy = 1)),
    vjust = -0.5,
    size = 5
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = NULL,
    y = "Accuracy",
    title = "Same model, different context",
    subtitle = "The helper model sees the plot correctly; the primary model ignores it"
  )
```

<!--

```{r mitm-helper-heatmap}
#| fig-width: 9
#| fig-height: 5
#| fig-alt: "Heatmap showing MITM helper accuracy per sample across epochs. The helper model correctly interprets most samples consistently."
mitm_helper |>
  filter(type == "mocked") |>
  plot_heatmap() +
  labs(
    title = "MITM helper accuracy by sample",
    subtitle = "Without prior-inducing context, the model sees the plots correctly"
  )
```

-->

<!--

### The diamonds example

We reversed the carat-price relationship in the `diamonds` dataset. Here's what the MITM helper reported:

```{r diamonds-helper}
diamonds_helper <- mitm_helper |>
  filter(id == "diamonds_carat_price_reversal") |>
  slice_head(n = 1) |>
  mutate(
    answer = str_remove(answer, "The description of the image content above.*")
  )
```

> `r diamonds_helper$answer`

The helper correctly describes a *negative* relationship. But the primary model reported a positive correlation anyway.

--> 

## What this means for now

